title: "MM20230322_summary_CFU"
author: "Mela3ie Mendel"
date: "23-3-2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# clean environment
```{r}
rm(list=ls())
```

```{r cars}
library(ggplot2)
library(dplyr)
library(multcompView)
library(agricolae)
library(ggsignif)
library(ggpubr)
library("report")
setwd("C:/Users/Mende012/OneDrive - Universiteit Utrecht/Projects/4. Pst effector library/4.14 Repeat test infiltration different strains/CFU")
```


# filter for samples of interest 
```{r}
pst_raw <- read.delim("LDM20230123_CFU_count_calculations_different_Pseudomonas.txt")
head(pst_raw)

#filter out all samples at t2
#pst_raw <- pst_raw %>% filter(dpi =="2")
#pst_raw[is.na(pst_raw)] <- 0
#pst_raw
```

# Pseudomonas strains as factors
```{r}
pst_raw$Pseudomonas.strain <- as.factor(pst_raw$Pseudomonas.strain)
pst_raw$Date <- as.factor(pst_raw$Date)

# remove rows with NA
pst_raw <- pst_raw[1:210,]
```

# log10 transformation of data for better presentation
```{r}
MIN <- min(pst_raw$CFU.ml..1cm.2[pst_raw$CFU.ml..1cm.2!=0]) # set lowest CFU counts as 0
pst_raw$CFU.log10 <- log10(pst_raw$CFU.ml..1cm.2+MIN)
```

# make df containing only data I want to plot: Pst strain, CFU log10 transormed

```{r}
Pst_strain <- as.factor(pst_raw$Pseudomonas.strain)
Pst_CFU <- pst_raw$CFU.log10
Pst_exp_date <- as.factor(pst_raw$Date)
Pst_OD <- pst_raw$OD6000.infiltrated.0dpi
Pst_comment <- pst_raw$Comment

simple_df <- data.frame(Pst_strain, Pst_OD, Pst_exp_date, Pst_CFU, Pst_comment)

simple_df_un <- simple_df[(simple_df$Pst_comment == "uninfiltrated"),]
simple_df <- simple_df[!(simple_df$Pst_comment == "uninfiltrated"),]
```


# Basic Box Plots comparing Pst strains
```{r}
# subset into different OD
df_OD2 <- simple_df[(simple_df$Pst_OD == "0.200"),]
df_OD4 <- simple_df[(simple_df$Pst_OD == "0.400"),]
df_OD6 <- simple_df[(simple_df$Pst_OD == "0.600"),]


# basic boxplot OD 0.2
 OD2 <- ggplot(df_OD2, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(4, 8.5)

OD2


# basic boxplot OD 0.4
 OD4 <- ggplot(df_OD4, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(4, 8.5)

OD4


# basic boxplot OD 0.6
 OD6 <- ggplot(df_OD6, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(4, 8.5)

OD6

ggarrange(OD2, OD4, OD6, ncol = 3)
```
# Plot by strain over time
```{r}
# subset simple df by strain
simple_df <- simple_df[!(simple_df$Pst_OD == "c"),]

Eth <- simple_df[(simple_df$Pst_strain == "EtHAn"),]
DC <- simple_df[(simple_df$Pst_strain == "DC3000"),]
D28 <- simple_df[(simple_df$Pst_strain == "D28E"),]
D36 <- simple_df[(simple_df$Pst_strain == "D36E"),]
CEL <- simple_df[(simple_df$Pst_strain == "deltaCEL"),]
hrc <- simple_df[(simple_df$Pst_strain == "hrcC"),]
psp <- simple_df[(simple_df$Pst_strain == "PSP"),]
```


```{r}
# 1. remove C from df
# 2. subset simple_df by bacteria
# 3. for each strain make a plot showing CFU counts of all 3 concentration per day
co <- c("#E69F00", "#009E73")

plot <- ggplot(psp, aes(x=as.factor(Pst_OD), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")              
               outlier.colour = "red",           # Outliers color, 
               alpha=0.9)+                    # Box color transparency
  labs(x= (unique(Pst_strain)), y = "log10 (CFU/cm2)") +
  

  geom_point(position = position_jitterdodge(0.1)) +
  theme_classic() +
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  scale_fill_manual(values = co) +
  scale_color_manual(values = co) +
  ylim(4, 9)

ggsave("MM20230403_psp.svg", plot, width = 3, height = 4, units = "in", dpi = 300)
```

# stats t-test
```{r}
all_OD <- as.vector(unique(simple_df$Pst_OD))
all_strain <- c(Eth, DC, D28, D36, CEL, hrc, psp)

tt <- function(df_Pst, OD, date1, date2){
a <- as.data.frame(df_Pst[(df_Pst$Pst_OD == OD),])
gr1 <- a[(a$Pst_exp_date == date1),]
gr2 <- a[(a$Pst_exp_date == date2),]
b <- t.test(gr1$Pst_CFU, gr2$Pst_CFU, alternative = "two.sided", var.equal = TRUE)
return(b$p.value)
}

# apply function over OD
for(ODs in all_OD){
  pval <- tt(df_Pst = psp, OD = ODs, date1 = "19-1-2023", date2 = "21-1-2023")
  print(pval)
} %>% unlist()

ts_Eth <- data.frame(all_OD, unlist())
  
# df to store results of t-test
df_t_res <- data.frame(all_OD, print(format(round(unlist(res_t),3), nsmall = 3))) 

# rename columns in a sensible way
colnames(df_t_res)[1] = "dpi"
colnames(df_t_res)[2] = "p-value"
```




```{r}
# write a function for the plot tested above
```{r}
# define the function 
plot_template <- function(data){
  ggplot(data, aes(x=as.factor(Pst_OD), y=Pst_CFU, color = Pst_exp_date))+
  
    geom_boxplot(outlier.colour = "red",           # Outliers color, 
                 alpha=0.9)+                    # Box color transparency
    labs(x= as.character(unique(Pst_strain)), y = "log10 (CFU/cm2)") +
  
    geom_point(position = position_jitterdodge(0.1)) +
    theme_classic() +
    guides(x = guide_axis(angle = 45 ))  +          # axis label
    scale_fill_manual(values = co) +
    scale_color_manual(values = co) +
    ylim(4, 9)
    }



# define the vector whichs element will be used in the plot_templat
strains <- unique(simple_df$Pst_strain)

subset <- subset(simple_df, select = Pst_strain)
# chatGPT beta

# Create a data frame with columns for country, gender, and age
df <- data.frame(country = c("USA", "Canada", "USA", "Canada"),
                 gender = c("Male", "Male", "Female", "Female"),
                 age = c(25, 30, 35, 40))

# Use the by function to split the data by country and gender:
plots <- by(simple_df, simple_df$Pst_strain, function(subset) {
  
  # Create a ggplot2 plot of age versus gender
  ggplot(subset, aes(x = gender, y = age)) +
    geom_bar(stat = "identity")
})

# Print each plot
for (i in seq_along(plots)) {
  print(plots[[i]])
}

```



# Basic Box Plots comparing Pst strains infiltrate
```{r}
# subset into different OD
df_OD2u <- simple_df_un[(simple_df_un$Pst_OD == "0.200"),]
df_OD4u <- simple_df_un[(simple_df_un$Pst_OD == "0.400"),]
df_OD6u <- simple_df_un[(simple_df_un$Pst_OD == "0.600"),]


# basic boxplot OD 0.2
 OD2u <- ggplot(df_OD2u, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(7, 9)

OD2u


# basic boxplot OD 0.4
 OD4u <- ggplot(df_OD4u, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(7, 9)

OD4u


# basic boxplot OD 0.6
 OD6u <- ggplot(df_OD2u, aes(x=as.factor(Pst_strain), y=Pst_CFU, color = Pst_exp_date)) +
  geom_boxplot(#fill= c("seagreen4", "red")                # box colour
               outlier.colour = NA,           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  labs(x= "Pseudomonas strain", y = "log10 (CFU/cm2)") +
  
  geom_point(position = position_jitterdodge(0.1)) +
  
  guides(x = guide_axis(angle = 45 ))  +          # axis label
  theme_classic() +
  ylim(4, 8.5)

OD6u

ggarrange(OD2u, OD4u, OD6u, ncol = 3)

```

# ANOVA
# 1. ANOVA to analyse wheather there are differences between the groups
# H0 All means are the same 
# H1 at least one mean is different

anova_pst <- aov(Pst_CFU ~ Pst_strain, data = simple_df)

  {summary(anova_pst)
    summary.lm(anova_pst)}

# Posthoc

```{r}
# 2. Posthoc to see which groups are significantly different - generate labels

# posthoc
  TUKEY <- TukeyHSD(anova_pst)
  
  generate_label_df <- function(TUKEY, variable){
    
    # Extract labels and factor levels from Tukey post-hoc
    Tukey.levels <- TUKEY[[variable]][,4]
    Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])
    
    #I need to put the labels in the same order as in the boxplot :
    Tukey.labels$Pst_strain = rownames(Tukey.labels)
    Tukey.labels=Tukey.labels[order(Tukey.labels$Pst_strain) , ]
    return(Tukey.labels)
    }

# Apply the function on my dataset

  LABELS <- generate_label_df(TUKEY , "Pst_strain")
  
  names(LABELS) <- c("Letters","Pst_strain")
  
  yvalue <- aggregate(Pst_CFU ~ Pst_strain, data = simple_df, mean) # obtain letter position for y axis using means
  
  LABELS$Pst_strain <- (unlist(lapply(strsplit(LABELS$Pst_strain,'[_]'),function(x){x[[1]]})))
  
  final <- merge(yvalue,LABELS) 
  
  stat_labels <- as.vector(LABELS$Letters) # extract letter column as a vector to use for plot
  
```

# add labels of TUKEY test
```{r}

# basic boxplot
 ggplot(simple_df, aes(x=as.factor(Pst_strain), y=Pst_CFU)) +
  geom_boxplot(fill="seagreen4",                 #box colour
               outlier.colour = "red",           # Outliers color, 
               alpha=0.9)+                       # Box color transparency
  xlab("Pseudomonas strain") +                   # axis label
  ylab("log10 (CFU/cm2)") +

   geom_jitter(shape=16, position=position_jitter(0.1)) +
   stat_boxplot(geom = "errorbar",               # Error bars
                    width = 0.2)
```


# check for batch effect

# select right data 
```{r}
selected <- c ("D36", "DC3000")
control_only <- simple_df[which(simple_df$Pst_strain == selected),] 
control_only

D36_only <- simple_df[which(simple_df$Pst_strain == "D36"),] 
D36_only

DC3000_only <- simple_df[which(simple_df$Pst_strain == "DC3000"),] 
DC3000_only
```





















